---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

# bugsigdbr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check-bioc](https://github.com/waldronlab/bugsigdbr/workflows/R-CMD-check-bioc/badge.svg)](https://github.com/waldronlab/bugsigdbr/actions)
[![Codecov test coverage](https://codecov.io/gh/waldronlab/bugsigdbr/branch/main/graph/badge.svg)](https://codecov.io/gh/waldronlab/bugsigdbr?branch=main)
<!-- badges: end -->

Utilities for accessing data from BugSigDB.org and creating plain text
signatures.

## Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `bugsigdbr` using from [Bioconductor](http://bioconductor.org/) the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("bugsigdbr")
```

And the development version from [GitHub](https://github.com/waldronlab/bugsigdbr) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("waldronlab/bugsigdbr")
```
## Example

```{r example, eval = requireNamespace('bugsigdbr')}
library("bugsigdbr")

bsdb <- bugsigdbr::importBugSigDB()
```

## Citation

Below is the citation output from using `citation('bugsigdbr')` in R. Please
run this yourself to check for any updates on how to cite __bugsigdbr__.

```{r 'citation', eval = requireNamespace('bugsigdbr')}
print(citation('bugsigdbr'), bibtex = TRUE)
```

Please note that the `bugsigdbr` was only made possible thanks to many other R and bioinformatics software authors, which are cited either in the vignettes and/or the paper(s) describing this package.

## Code of Conduct

Please note that the `bugsigdbr` project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.

This package was developed using `r BiocStyle::Biocpkg('biocthis')`.

