---
title: "R-side access to published microbial signatures from BugSigDB"
author: "Ludwig Geistlinger, Jennifer Wokaty, and Levi Waldron"
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('bugsigdbr')`"
vignette: >
  %\VignetteIndexEntry{R-side access to BugSigDB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

# BugSigDB: a comprehensive database of published microbial signatures 

[BugSigDB](https://bugsigdb.org) is a manually curated database of microbial 
signatures from the published literature of differential abundance studies of 
human and other host microbiomes.

BugSigDB provides:

* standardized data on geography, health outcomes, host body sites, and 
  experimental, epidemiological, and statistical methods using controlled vocabulary,
* results on microbial diversity,
* microbial signatures standardized to the NCBI taxonomy, and
* identification of published signatures where a microbe has been reported.

The [bugsigdbr](https://github.com/waldronlab/bugsigdbr) package implements 
convenient access to BugSigDB from within R/Bioconductor. 
The goal of the package is to facilitate import of BugSigDB data into R/Bioconductor,
provide utilities for extracting microbe signatures, and enable export of the 
extracted signatures to plain text files in standard file formats such as 
[GMT](https://software.broadinstitute.org/cancer/software/gsea/wiki/index.php/Data_formats#GMT:_Gene_Matrix_Transposed_file_format_.28.2A.gmt.29).

The [bugsigdbr](https://github.com/waldronlab/bugsigdbr) package is primarily a
data package. For descriptive statistics and comprehensive analysis of BugSigDB
contents, please see the 
[BugSigDBStats package](https://github.com/waldronlab/BugSigDBStats)
and 
[analysis vignette](http://waldronlab.io/BugSigDBStats/articles/BugSigDBStats.html).
 
We start by loading the package

```{r "start", message=FALSE}
library(bugsigdbr)
```

## Obtaining published microbial signatures from BugSigDB

```{r getBugSigDB}
bsdb <- importBugSigDB()
colnames(bsdb)
```

## Extracting microbe signatures

```{r getSignatures}
sigs <- getSignatures(bsdb)
length(sigs)
sigs[1:3]
```

## Writing microbe signatures to file in GMT format

```{r writeGMT}
writeGMT(sigs, gmt.file = "bugsigdb_signatures.gmt")
```

```{r, echo = FALSE}
file.remove("bugsigdb_signatures.gmt")
```

## Displaying BugSigDB signature and taxon pages

```{r browse}
browseSignature(names(sigs)[1])
browseTaxon(sigs[[1]][1])
```

# Session info

```{r sessionInfo}
sessionInfo()
```
